---
title: "Global Watershed Database Compilation Notebook"
output: html_notebook
---

# ---- SETUP ----

Load packages, run reference scripts, and define settings:
```{r}
library(dplyr)
library(tidyverse)
library(data.table)
library(sf)
library(rnaturalearth)
library(terra)
library(mapview)
library(gert)

source("ref_scripts/watersheds_functions.R")
source("ref_scripts/general_functions.R")

options(scipen = 999) # Avoid scientific notation in number formatting
```


# ---- FIRST BASIN LAYER ----

This is the broadest, least granular basin layer. 
We use the the Basins and Sub-Basins dataset from the World Meteorological Organization's (WMO) Global Runoff Data Center (GRDC). 
The WMO layer is partly derived from the HydroBASINS ("hybas") layer from HydroSHEDS, but it is simpler and contains proper names for all basins and sub-basins.
We merge WMO polygons into basin-level geometries for this layer, composite this with the HydroBASINS level 1 layer, and keep features from both to be thorough. 

Read in and initially clean raw data:
```{r}
# Read all regional hybas level 1 layers and merge into global layer
hybas_glob_comb_l1 <- hybas_read_multi("watersheds_data/HydroBASINS", "hybas_lake_.*_lev01-12_v1c$", "lev01_v1c.shp$")

# Read in WMO layer, clean up useful variables, and validate geometries
wmo_basins <- st_read("watersheds_data/wmobb_json/wmobb_basins.json") %>%
  mutate(across(all_of(c("WMOBB_NAME", "WMOBB_BASIN", "WMOBB_SUBBASIN")), ~ title_case(na_if(str_remove(., " \\(.*$"), "---"))), # Convert from all caps to title case
         WMOBB_DESCRIPTION = paste0(toupper(substr(WMOBB_DESCRIPTION, 1, 1)), substr(WMOBB_DESCRIPTION, 2, nchar(WMOBB_DESCRIPTION)))) %>% # Only capitalize first letter
  st_make_valid() %>%
  st_wrap_dateline(options = c("WRAPDATELINE=YES", "DATELINEOFFSET=180")) # Properly split geometries across 0/180 degree dateline if needed
```

Create merged basin-level WMO layer additionally split and indexed by hybas-derived regions and basin divisions, if applicable:
```{r}
# Intersect WMO layer with hybas level 1 layer to assign common Region variable between both layers 
wmo_ovl1 <- st_intersection(wmo_basins, select(hybas_glob_comb_l1, Region)) %>% 
  st_make_valid()

# Extract any WMO features that do not intersect with hybas and assign Region as "Other"
wmo_int_matrix <- st_intersects(wmo_basins, hybas_glob_comb_l1, sparse = FALSE)
wmo_unm1 <- wmo_basins[rowSums(wmo_int_matrix) == 0, ] %>%
  mutate(Region = "Other")

# Combine all WMO intersecting and non-intersecting features into unified layer and merge geometry grouped by unique basins and regions
wmo_comb_l1 <- bind_rows(wmo_ovl1, wmo_unm1) %>%
  filter(!st_is_empty(geometry), st_geometry_type(geometry) %in% c("POLYGON", "MULTIPOLYGON")) %>%
  mutate(geometry = st_cast(geometry, "MULTIPOLYGON")) %>%
  group_by(WMOBB_BASIN, REGNAME, Region) %>%
  summarise(geometry = st_union(geometry), .groups = "drop") %>%
  st_make_valid()
```

Modify basin-level hybas layer (level 1) to be additionally split and indexed by WMO-derived regions and basin divisions, if applicable:
```{r}
# Intersect hybas_glob_comb_l1 with wmo_ovl1 to assign subdivisions
hybas_ovl_wmo1 <- st_intersection(select(wmo_ovl1, WMOBB_BASIN, REGNAME), hybas_glob_comb_l1)

# Find basin features that do NOT intersect any WMO feature
hybas_unm_wmo1 <- hybas_glob_comb_l1 %>%
  st_make_valid() %>%
  filter(!lengths(st_intersects(., wmo_basins)) > 0)  # fix: ensure logical vector

# Combine basin intersecting and non-intersecting parts
hybas_comb_wmo1 <- bind_rows(hybas_ovl_wmo1, hybas_unm_wmo1) %>%
  filter(!st_is_empty(geometry), st_geometry_type(geometry) %in% c("POLYGON", "MULTIPOLYGON")) %>%
  mutate(geometry = st_cast(geometry, "MULTIPOLYGON")) %>%
  group_by(Region, WMOBB_BASIN, REGNAME) %>%
  summarise(geometry = st_union(geometry), .groups = "drop") %>%
  st_make_valid()

# Combine the two layers and drop duplicates based on identifier combo
basins_comb_l1 <- bind_rows(wmo_comb_l1, hybas_comb_wmo1) %>%
  distinct(Region, WMOBB_BASIN, REGNAME, .keep_all = TRUE) %>%
  filter(st_area(.) >= units::set_units(1000000, "m^2"))
```


# ---- SECOND BASIN LAYER ----

Composite WMO sub-basin-level layer with hybas level 1 layer
```{r}
# WMO Level 2 version — keep finest granularity
wmo_comb_l2 <- bind_rows(wmo_ovl1, wmo_unm1) %>%
  filter(!st_is_empty(geometry), st_geometry_type(geometry) %in% c("POLYGON", "MULTIPOLYGON")) %>%
  mutate(geometry = st_cast(geometry, "MULTIPOLYGON")) %>%
  select(WMOBB:REGNAME, WMO306_MoC_REFERENCE, Region) %>%
  st_make_valid()

hybas_unm_wmo2 <- hybas_unm_wmo1 %>%
  mutate(
    WMOBB = NA_integer_,
    WMOBB_NAME = NA_character_,
    WMOBB_BASIN = NA_character_,
    WMOBB_SUBBASIN = NA_character_,
    WMOBB_DESCRIPTION = NA_character_,
    REGNAME = NA_character_,
    WMO306_MoC_REFERENCE = NA_character_
  ) %>%
  select(WMOBB:REGNAME, WMO306_MoC_REFERENCE, Region, geometry)

basins_comb_l2 <- bind_rows(wmo_comb_l2, hybas_unm_wmo2) %>%
  filter(st_area(.) >= units::set_units(1000000, "m^2"))
```


# ---- THIRD BASIN LAYER ----

Composite second basin layer with global hybas level 5 layer:
```{r}
hybas_glob_comb_l5 <- hybas_read_multi("/Volumes/OWC Elite Pro Dual/Data + Other/Minnow/watersheds_data/HydroBASINS", "hybas_lake_.*_lev01-12_v1c$", "lev05_v1c.shp$")

# Get shared Region values
region_groups <- intersect(
  unique(hybas_glob_comb_l5$Region),
  unique(basins_comb_l2$Region)
)

# Apply st_intersection region by region
hybas_l5_wmo_comp <- purrr::map_dfr(region_groups, function(rg) {
  
  hybas_sub <- hybas_glob_comb_l5 %>%
    filter(Region == rg) %>%
    select(HYBAS_ID)
  
  basins_sub <- basins_comb_l2 %>%
    filter(Region == rg) %>%
    select(WMOBB:Region)
  
  st_intersection(hybas_sub, basins_sub)
}) %>%
  filter(st_area(.) >= units::set_units(1000000, "m^2"))
```


# ---- FOURTH BASIN LAYER ----

Read and merge global hybas level 6 data into one layer:
```{r}
hybas_glob_comb_l6 <- hybas_read_multi("/Volumes/OWC Elite Pro Dual/Data + Other/Minnow/watersheds_data/HydroBASINS", "hybas_lake_.*_lev01-12_v1c$", "lev06_v1c.shp$")
```


# ---- EXPORT ALL BASIN LAYERS ----

```{r}
basins_comb_l1_file <- "watersheds_data/global_basins_wmo_hybas1_composite_layer_1.gpkg"
if (file.exists(basins_comb_l1_file)) file.remove(basins_comb_l1_file)
st_write(basins_comb_l1, basins_comb_l1_file)

basins_comb_l2_file <- "watersheds_data/global_basins_wmo_hybas1_composite_layer_2.gpkg"
if (file.exists(basins_comb_l2_file)) file.remove(basins_comb_l2_file)
st_write(basins_comb_l2, basins_comb_l2_file)

hybas_l5_wmo_comp_file <- "watersheds_data/global_basins_wmo_hybas5_composite_layer_3.gpkg"
if (file.exists(hybas_l5_wmo_comp_file)) file.remove(hybas_l5_wmo_comp_file)
st_write(hybas_l5_wmo_comp, hybas_l5_wmo_comp_file)

hybas_glob_comb_l6_file <- "watersheds_data/global_basins_hybas6_layer_4.gpkg"
if (file.exists(hybas_glob_comb_l6_file)) file.remove(hybas_glob_comb_l6_file)
st_write(hybas_glob_comb_l6, hybas_glob_comb_l6_file)
```


# ---- READ, SIMPLIFY, AND INDEX RIVER LAYERS ----

```{r}
rivers_glob <- st_read("watersheds_data/HydroRIVERS/HydroRIVERS_v10_shp/HydroRIVERS_v10_shp/HydroRIVERS_v10.shp")

rivers_low <- filter(rivers_glob, ORD_FLOW <= 3 | ORD_STRA >= 6) %>%
  group_by(ORD_FLOW) %>%
  summarise(geometry = st_union(geometry), .groups = "drop") %>%
  st_cast("MULTILINESTRING") %>%
  st_make_valid() %>%
  st_intersection(basins_comb_l1)

rivers_mod <- filter(rivers_glob, ORD_FLOW <= 4 | ORD_STRA >= 5) %>%
  group_by(ORD_FLOW) %>%
  summarise(geometry = st_union(geometry), .groups = "drop") %>%
  st_cast("MULTILINESTRING") %>%
  st_intersection(basins_comb_l2)

rivers_all <- rivers_glob %>%
  group_by(ORD_FLOW) %>%
  summarise(geometry = st_union(geometry), .groups = "drop") %>%
  st_cast("MULTILINESTRING") %>%
  st_intersection(hybas_l5_wmo_comp)# %>%
#  st_intersection(select(hybas_glob_comb_l6, HYBAS_ID, geometry))
```

